Writing Trash Bin

### 2024-03-11 paper.qmd
finished uncertainty is inference section

# What is Uncertainty?
## Colloquial Definitions
Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.

@@

## Taxonomy Definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7*). In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertianty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

## Defining uncertainty with respect to inference
What exactly is uncertainty, then? If we were to consider making this overarching definition, what would it need in order to be "encapsulating"? Well, let us consider some comments about uncertainty or common sentiments that come up time and time again in the literature.

We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?

@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field.
For example, do descriptive statistics have uncertainty? Descriptive statistics simply describe features of your sample as they are. The sample mean is the sample mean, the sample size is the sample size, and the sample variance is its sample variance. We would not say we are "uncertain" about these values unless we are questioning our own computation skills.

Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics. The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis. At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023]. When we move onto data provenance the issue persists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014]. After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023]. Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014]. These cases highlight that uncertainty is defined at *every* stage in relation to our motivating question, from data collection to visualisation. Discussions of uncertainty cannot be had if we are not clear *what* we are uncertain about. Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.

Variance and probability are typically considered the closest of these concepts to uncertainty, however they clearly do not fully encapsulate the concept. @Spiegelhalter2017 makes the reason for this distinction clear, as in the absence of something to draw inference on, variance and probability describe precise random events (such as a coin flip), while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased). A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty.

>>>>

# What is Uncertainty?
## Colloquial definitions
Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.

@@

## Taxonomy definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7*). In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertianty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

## Defining uncertainty with respect to inference
What exactly is uncertainty, then? If we were to consider making this overarching definition, what would it need in order to be "encapsulating"? Well, let us consider what might *not* be considered uncertain in order to understand this concept a little better.

@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field and in doing so, highlights an interesting connection between inference and uncertainty. Descriptive statistics simply describe our sample as it is and summarises large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however this was not always the case. Around the 19th century in England, *posivitism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023]. By its very nature, descriptive statistics *cannot* be used to make inferences about the data because it simply exists to summaries the data, to use it to make statements about *new* data is incorrect useage. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e. that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023]. This subtle shift, from descriptive statistics to inferential statistics was historically shunned *due to the fact it introduced the unknowable*, or in other words, uncertainty.

This philosophical understanding of statistics highlights that descriptive statistics do not have uncertainty, however some readers may disagreee with that statement. Variance and probability are typically considered stand ins for "uncertainty", it is often how we choose to measure it, and since probability and variance exist in descriptive statistics, descriptive statistics *must* have uncertainty. This is not necessarily true, and related distinction was made by @Spiegelhalter2017  commented on a differentiation between precise random events (such as the probability of a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased). A sample variance is not unknown, and therefore it is not uncertain, rather it is a precise description of dispersion. If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the the variance would become relevant in our discussions of uncertainty, but in isolation it is not uncertain.

The idea that inference is related to uncertainty pops up in most discussions of uncertainty, however, due to the now *implicit* understanding we are performing inferential statistics, it is often brought up as a *task* or *goal* dependence. This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e. the task or goal. The task or goal is, in essence, the *statistic you are drawing inference on*. Were this not true, we would have no reason to differentiate between a prediction interval and a sampling distribution because they could both be considered "uncertainty about the mean", however they are different because their inferential goals are different.

These examples highlight two key concepts that seem to be true about the relationship between uncertainty and inference:    
1) Uncertainty is the by-product of inference, as we seek to draw conclusions about something unknowable.    
2) Uncertainty must be defined with respect to a *specific* statistic we wish to draw inference on, there is no such thing as a *general* uncertainty that is not linked to a motivating question.

## Mistakes when we misunderstand uncertainty

### 2024-03-11 writingdumpster.qmd
finished uncertainty is inference section

# Main section outtakes
- What exactly is uncertainty, then? Let us consider the concepts related to uncertainty that have been discussed so far. We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics. For example, I wouldn't describe the population parameters of a distribution I have simulated to be "uncertain"

>>>>

# Main section outtakes
- This is likely because The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics. The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
- We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?
- What exactly is uncertainty, then? Let us consider the concepts related to uncertainty that have been discussed so far. We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics. For example, I wouldn't describe the population parameters of a distribution I have simulated to be "uncertain"
